Squintability and Other Metrics for Assessing Projection Pursuit Indexes, and Guiding Optimisation Choices

### 2025-03-02 jso.qmd
updating description of simulations

This section describes the details of two simulation studies. The first simulation
compares the JSO performance with an existing optimiser, Creeping Random
Search (CRS) [@RJ-2021-105; @laa_using_2020] to explore JSO's behaviour
under different data dimensions. The second simulation examines various factors affecting the performance of JSO.
The JSO performance is evaluated under different hyper-parameters combinations
(number of jellyfish and the maximum number of tries). The effect of index properties (smoothness and squintability) along
with JSO hyper-parameters and data dimension variations is captured and studied using logistic regression.

The performance of JSO is measured by the success rate, defined as the
proportion of simulations that achieves a final index value within 0.05
of the best index value found among all 50 simulations (see
@fig-success-rate for an illustration). This comparison is based on
projection pursuit to find the pipe shape investigated by
@laa_using_2020 using the `holes` index.

@@

The JSO performance is compared to the existing optimizer, CRS, in a pipe-finding problem using the holes index in different dimensional spaces. Fifty simulations are conducted in four data dimensions ($d = 6, 8, 10, 12$) with each optimiser. JSO uses
100 jellyfish with a maximum of 100 tries, while the CRS allows a
maximum of 1000 samples at each iteration before the algorithm
terminates. The different numbers account for the multiple paths of JSO
to enable fairer comparison with CRS. The results of the simulation are
collected using the data structure proposed in @RJ-2021-105 for
assessing JSO, where the design parameters are stored along with index
value, projection basis, random seed, and computation time.

```{r}
#| label: fig-success-rate
#| fig.cap: "Illustration of success rate calculation: Final projections based on projection pursuit to find the pipe shape in 8D data using the holes index, optimised by CRS, in 50 simulations. The 50 final projections are sorted by their index values. The highest index value found across all simulations is 0.969. Out of the 50 simulations, 43 achieved an index value within 0.05 of the best, resulting in a success rate of 0.86 (43/50)."
# pipe_jellyfish_12d <- pipe_better |> 
#   filter(dim == 2) |> 
#   mutate(dim = as.numeric(dim) * 8, sim = as.numeric(sim)) 
# proj_dt <- map_dfr(pipe_jellyfish_12d |> pull(basis),
#               ~as_tibble(pipe1000_8d %*% .x), .id = "sim") |>
#   mutate(sim = as.numeric(sim)) |>
#   left_join(pipe_jellyfish_12d |> select(sim, index_val), by = "sim")
# 
# idx_val_sorted <- proj_dt |> pull(index_val) |> unique() |> sort()
# proj_dt |>
#   ggplot(aes(x = V1, y = V2)) +
#   geom_point(size = 0.5) +
#   geom_text(aes(label = round(index_val, 3)), x = 0, y = 3.7,
#             size = 8) +
#   xlim(-4, 4) + ylim(-4, 4) +
#   facet_wrap(~fct_reorder(as.factor(sim), -index_val), ncol = 10) +
#   theme_bw() +
#   theme(aspect.ratio = 1, axis.ticks = element_blank(),
#         axis.text = element_blank(), axis.title = element_blank(),
#         panel.grid = element_blank(), strip.background = element_blank(),
#         strip.text = element_blank()) 
knitr::include_graphics(here::here("figures", "success-rate.png"))
```

@@

To examine the performance of JSO across various hyper-parameter
combinations, the pipe-finding problem using the holes index is repeated
for fifty times in each hyper-parameter combination: 20, 50, and 100
jellyfish and a maximum of 50 and 100 attempts. The success rate is
calculated for each combination.

@@

To construct a relationship among success rate, smoothness,
squintability, and JSO hyper-parameters, a generalised linear model is
fitted using a binomial family and a logit link function. The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) <!-- creating a new binary variable `long_time` to
indicate cases with an average run time over 20 seconds, and 3)-->
re-coding the success rate for the `stringy` index as 0, because none of
the 50 simulations correctly identified the sine-wave shape.

@@

```{r}
#| label: fig-proportion
#| fig.width: 8
#| fig.height: 3.5
#| fig.align: "center"
#| fig.cap: "The proportion of simulations that reach near-optimal index values in the pipe-finding problem using the holes index. The proportion is calculated based on the number of simulations, out of 50, that achieve an index value within 0.05 of the best-performing simulation. To quantify uncertainty, 500 bootstrap samples are generated. The thin segments represent the proportion for each of the 500 bootstrap replicates, while the thicker segments represent the mean of these bootstrap samples, connnected by lines. As the dimensionality increases, the proportion of simulations reaching the optimal index value decreases."
pipe_sim_best <- sim_pipe_run_best |> 
  group_by(n_jellies, max_tries, d) |>
  summarise(I_best = max(I_max))

>>>>

This section explains the details of two simulation studies: (1) comparison between the existing Creeping Random Search (CRS) [@RJ-2021-105; @laa_using_2020] and JSO, and (2) examining the factors affecting the performance of JSO.

The second simulation examines various factors affecting the performance of JSO.
The JSO performance is evaluated under different hyper-parameters combinations
(number of jellyfish and the maximum number of tries). The effect of index properties (smoothness and squintability) along
with JSO hyper-parameters and data dimension variations is captured and studied using logistic regression.

@@

The CRS is the main optimisation routine currently used for the guided tour. Here the two optimisers are compared on the task of finding the pipe using the `holes` index in data with dimensions, $d = 6, 8, 10, 12$. Fifty simulated data sets are used for each dimension. JSO uses 100 jellyfish with a maximum of 100 tries, while the CRS allows a maximum of 1000 tries at each iteration before the algorithm terminates. These choices enable fair comparison between CRS and JSO, that conforms to how they are used in practice.

The performance of the optmisers is measured by the success rate, which is defined as the proportion of simulations that achieves a final index value within 0.05
of the best index value found among all 50 simulations. @fig-success-rate illustrates why the choice of 0.05 is reasonable: the pipe is not recognisable in the projected data. This is motivated by @laa_using_2020's approach to investigating PP indexes.

The results of the simulation are collected using the data structure used in @RJ-2021-105 for assessing PP optimisers. The design parameters are stored along with index value, projection basis, random seed, and computation time.

```{r}
#| label: fig-success-rate
#| fig.cap: "How success rate is calculated, illustrated using the optimal projections from 50 optimisations of 8D pipe data sorted by index value. The pipe shape is recognisable in the projection index values between 0.933-0.969. Of the 50 simulations, 43 achieved an index value within 0.05 of the best, resulting in a success rate of 0.86."
# pipe_jellyfish_12d <- pipe_better |> 
#   filter(dim == 2) |> 
#   mutate(dim = as.numeric(dim) * 8, sim = as.numeric(sim)) 
# proj_dt <- map_dfr(pipe_jellyfish_12d |> pull(basis),
#               ~as_tibble(pipe1000_8d %*% .x), .id = "sim") |>
#   mutate(sim = as.numeric(sim)) |>
#   left_join(pipe_jellyfish_12d |> select(sim, index_val), by = "sim")
# 
# idx_val_sorted <- proj_dt |> pull(index_val) |> unique() |> sort()
# proj_dt |>
#   ggplot(aes(x = V1, y = V2)) +
#   geom_point(size = 0.5) +
#   geom_text(aes(label = sprintf("%.3f", index_val)), x = 0, y = 3.7,
#             size = 8) +
#   xlim(-4, 4) + ylim(-4, 4) +
#   facet_wrap(~fct_reorder(as.factor(sim), -index_val), ncol = 10) +
#   theme_bw() +
#   theme(aspect.ratio = 1, axis.ticks = element_blank(),
#         axis.text = element_blank(), axis.title = element_blank(),
#         panel.grid = element_blank(), strip.background = element_blank(),
#         strip.text = element_blank()) 
knitr::include_graphics(here::here("figures", "success-rate.png"))
```

@@

To examine the performance of JSO across various hyper-parameter
combinations, the pipe-finding problem using the holes index is repeated
for fifty times in each hyper-parameter combination: 20, 50, and 100
jellyfish and a maximum of 50 and 100 attempts.

@@

A generalised linear model is fitted using a binomial family and a logit link function to assess the factors affecting success rate.  Predictors are smoothness,
squintability, and JSO hyper-parameters. A success rate of 0 indicates the optimiser did not come sufficiently close to the optimal projection. <!-- The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) creating a new binary variable `long_time` to
indicate cases with an average run time over 20 seconds, and 3)
re-coding the success rate for the `stringy` index as 0, because none of
the 50 simulations correctly identified the sine-wave shape.-->

@@

```{r}
#| label: fig-proportion
#| fig.width: 8
#| fig.height: 3.5
#| fig.align: "center"
#| fig.cap: "Summary of the relationship between JSO hyper-parameters on optimisation success, using the holes index for pipe data of dimensions 4-12. Bootstrap samples show the variability in success rate. Success rate mostly plateaus by 50 jellies. The JSO has some difficulty finding the pipe when dimension is higher than 8! Maximum number of tries has little effect."
pipe_sim_best <- sim_pipe_run_best |> 
  group_by(n_jellies, max_tries, d) |>
  summarise(I_best = max(I_max))

### 2025-03-02 jso.tex
updating description of simulations

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{anyfontsize}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{prop}{\protect\propositionname}
\providecommand{\definitionname}{Definition}
\providecommand{\propositionname}{Proposition}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, titlerule=0mm, coltitle=black, left=2mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, rightrule=.15mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toptitle=1mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomtitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6]

@@

This section describes the details of two simulation studies. The first
simulation compares the JSO performance with an existing optimiser,
Creeping Random Search (CRS) (Zhang et al. 2021; Laa and Cook 2020) to
explore JSO's behaviour under different data dimensions. The second
simulation examines various factors affecting the performance of JSO.
The JSO performance is evaluated under different hyper-parameters
combinations (number of jellyfish and the maximum number of tries). The
effect of index properties (smoothness and squintability) along with JSO
hyper-parameters and data dimension variations is captured and studied
using logistic regression.

The performance of JSO is measured by the success rate, defined as the
proportion of simulations that achieves a final index value within 0.05
of the best index value found among all 50 simulations (see
Figure~\ref{fig-success-rate} for an illustration). This comparison is
based on projection pursuit to find the pipe shape investigated by Laa
and Cook (2020) using the \texttt{holes} index.

@@

The JSO performance is compared to the existing optimizer, CRS, in a
pipe-finding problem using the holes index in different dimensional
spaces. Fifty simulations are conducted in four data dimensions
(\(d = 6, 8, 10, 12\)) with each optimiser. JSO uses 100 jellyfish with
a maximum of 100 tries, while the CRS allows a maximum of 1000 samples
at each iteration before the algorithm terminates. The different numbers
account for the multiple paths of JSO to enable fairer comparison with
CRS. The results of the simulation are collected using the data
structure proposed in Zhang et al. (2021) for assessing JSO, where the
design parameters are stored along with index value, projection basis,
random seed, and computation time.

@@

\caption{\label{fig-success-rate}Illustration of success rate
calculation: Final projections based on projection pursuit to find the
pipe shape in 8D data using the holes index, optimised by CRS, in 50
simulations. The 50 final projections are sorted by their index values.
The highest index value found across all simulations is 0.969. Out of
the 50 simulations, 43 achieved an index value within 0.05 of the best,
resulting in a success rate of 0.86 (43/50).}

@@

To construct a relationship among success rate, smoothness,
squintability, and JSO hyper-parameters, a generalised linear model is
fitted using a binomial family and a logit link function. The data is
pre-processed by 1) scaling the JSO hyper-parameters by a factor of 10
for interpretation, 2) re-coding the success rate for the
\texttt{stringy} index as 0, because none of the 50 simulations
correctly identified the sine-wave shape.

@@

\caption{\label{fig-proportion}The proportion of simulations that reach
near-optimal index values in the pipe-finding problem using the holes
index. The proportion is calculated based on the number of simulations,
out of 50, that achieve an index value within 0.05 of the
best-performing simulation. To quantify uncertainty, 500 bootstrap
samples are generated. The thin segments represent the proportion for
each of the 500 bootstrap replicates, while the thicker segments
represent the mean of these bootstrap samples, connnected by lines. As
the dimensionality increases, the proportion of simulations reaching the
optimal index value decreases.}

@@

\toprule
\textbf{Characteristic} & \textbf{OR} \textbf{(95\% CI)}\textsuperscript{\textit{1}} & \textbf{p-value} \\ 
\midrule\addlinespace[2.5pt]
Smoothness & 1.00 (0.90 to 1.11) & 0.928 \\ 
Squintability & 1.24 (1.13 to 1.39) & {\bfseries <0.001} \\ 
Dimension & 0.57 (0.44 to 0.73) & {\bfseries <0.001} \\ 
Number of jellyfish & 1.32 (1.17 to 1.50) & {\bfseries <0.001} \\ 
Maximum number of tries & 1.06 (0.94 to 1.21) & 0.333 \\ 
\bottomrule

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{orcidlink}
\usepackage{algorithm}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\theoremstyle{plain}
\newtheorem{defn}{\protect\definitionname}
\newtheorem{prop}{\protect\propositionname}
\providecommand{\definitionname}{Definition}
\providecommand{\propositionname}{Proposition}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, opacitybacktitle=0.6, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, opacityback=0, breakable, bottomrule=.15mm, bottomtitle=1mm, titlerule=0mm, coltitle=black, colframe=quarto-callout-note-color-frame, title={Algorithm: Jellyfish Optimizer Pseudo Code}, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, left=2mm]

@@

This section explains the details of two simulation studies: (1)
comparison between the existing Creeping Random Search (CRS) (Zhang et
al. 2021; Laa and Cook 2020) and JSO, and (2) examining the factors
affecting the performance of JSO.

The second simulation examines various factors affecting the performance
of JSO. The JSO performance is evaluated under different
hyper-parameters combinations (number of jellyfish and the maximum
number of tries). The effect of index properties (smoothness and
squintability) along with JSO hyper-parameters and data dimension
variations is captured and studied using logistic regression.

@@

The CRS is the main optimisation routine currently used for the guided
tour. Here the two optimisers are compared on the task of finding the
pipe using the \texttt{holes} index in data with dimensions,
\(d = 6, 8, 10, 12\). Fifty simulated data sets are used for each
dimension. JSO uses 100 jellyfish with a maximum of 100 tries, while the
CRS allows a maximum of 1000 tries at each iteration before the
algorithm terminates. These choices enable fair comparison between CRS
and JSO, that conforms to how they are used in practice.

The performance of the optmisers is measured by the success rate, which
is defined as the proportion of simulations that achieves a final index
value within 0.05 of the best index value found among all 50
simulations. Figure~\ref{fig-success-rate} illustrates why the choice of
0.05 is reasonable: the pipe is not recognisable in the projected data.
This is motivated by Laa and Cook (2020)'s approach to investigating PP
indexes.

The results of the simulation are collected using the data structure
used in Zhang et al. (2021) for assessing PP optimisers. The design
parameters are stored along with index value, projection basis, random
seed, and computation time.

@@

\caption{\label{fig-success-rate}How success rate is calculated,
illustrated using the optimal projections from 50 optimisations of 8D
pipe data sorted by index value. The pipe shape is recognisable in the
projection index values between 0.933-0.969. Of the 50 simulations, 43
achieved an index value within 0.05 of the best, resulting in a success
rate of 0.86.}

@@

A generalised linear model is fitted using a binomial family and a logit
link function to assess the factors affecting success rate. Predictors
are smoothness, squintability, and JSO hyper-parameters. A success rate
of 0 indicates the optimiser did not come sufficiently close to the
optimal projection.

@@

\caption{\label{fig-proportion}Summary of the relationship between JSO
hyper-parameters on optimisation success, using the holes index for pipe
data of dimensions 4-12. Bootstrap samples show the variability in
success rate. Success rate mostly plateaus by 50 jellies. The JSO has
some difficulty finding the pipe when dimension is higher than 8!
Maximum number of tries has little effect.}

@@

\toprule
\textbf{Characteristic} & \textbf{OR} \textbf{(95\% CI)}\textsuperscript{\textit{1}} & \textbf{p-value} \\ 
\midrule\addlinespace[2.5pt]
Smoothness & 1.00 (0.90 to 1.11) & 0.928 \\ 
Squintability & 1.24 (1.13 to 1.39) & \textbf{<0.001} \\ 
Dimension & 0.57 (0.44 to 0.73) & \textbf{<0.001} \\ 
Number of jellyfish & 1.32 (1.17 to 1.50) & \textbf{<0.001} \\ 
Maximum number of tries & 1.06 (0.94 to 1.21) & 0.333 \\ 
\bottomrule
